\Large \textbf{ Phenotype and gene ontology enrichment as guides for disease modeling in \cel{} }

### 2017-01-12 pea_paper_docs/tex/pea.tex
added figures to paper

% this is required to include graphics
\usepackage{graphicx}

@@

After we identified worm homologs for each trait, we reassessed how many traits
still had $>100$ gene candidates, and dropped all traits that had less than this
for our analysis. We identified 18 traits that met this criteria. The gene
lists for each of these 18 traits were then analyzed for gene, tissue and
phenotype enrichment. Tissue enrichment was performed using the WormBase Tissue
Enrichment Analysis tool (TEA)~\cite{}.

@@

Lupus-associated homologs were reasonably well annotated. Slightly more than half
of the genes had at least one phenotype annotation ($76/\wlupus{}$) and almost
all genes were annotated to at least one tissue or gene ontology term
($104/\wlupus{}$ and $115/\wlupus$ genes respectively).
We found that Lupus-associated homologs were enriched in `aneuploidy' (7 genes,
\qval{1}) and `meiotic chromosome segregation' (8 genes, \qval{1}). `Cell fate
transformation' (6 genes, \qval{1}), and `excess intestinal cells'
(5 genes, \qval{1}) were also overrepresented, as was `male tail morphology'
(6 genes, \qval{1}). Finally, the phenotype `nonsense mRNA accumulation' was also
enriched (5 genes, \qval{1}).
Meanwhile, TEA suggested that the
`excretory duct cell' (5 genes, \qval{2}) and the `posterior gonad arm' are
overrepresented in this dataset. We also found that the Pn.p cells P3.p through
P8.p were enriched in this dataset (5 genes, \qval{1}). GO enrichment pointed
at `modification-dependent macromolecule catabolic process' (23 genes,
\qval{15}) as a molecular function that characterizes this dataset. However,
this GO term was enriched only due to a single gene family, the \emph{skr} gene
family. Almost the entire \emph{skr} family was considered a candidate homolog
to the SKP1 human gene, making the GO enrichment suspect.

@@

\subsubsection*{Enrichment in the ciliary neuronal transcriptome}
As an example of how ontology enrichment can improve our understanding of
transcriptomes,
we selected a ciliary neuron dataset~\cite{} and ran the complete WormBase
Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
they are also present in the male cephalic sensillum and hermaphrodites also have
ciliated neurons. PEA reveals that the ciliated
neuron transcriptome is enriched for genes that are typically associated with
`meiotic chromosome segregation' (46 genes, \qval{5}),
`aneuploidy' (42 genes, \qval{5}) and `spindle defective early embryos' (45
genes, \qval{2}).

@@

We extracted all of the \cel{} genes that have been associated with an Egl
phenotype and we used TEA to understand what tissues are enriched. The HSN
was enriched more than five-fold above background (\qval{7}) as were vulD, vulC,
vulE and vulF (\qval{6}). The vulA, vulB2 and vulB1 were
enriched at slightly lower levels (\qval{5}), whereas the uterine muscles and
uterine seam cells were enriched more than twice above background levels
(\qval{2}).
Therefore, the Egl phenotype would seem to emerge primarily from defects in the
HSN, secondarily from defects in the vulva, and only sometimes from defects in the
uterine seam cells or muscles. It is notable that all vulval cells were not equally
enriched. Although all the `vul' cells are annotated to a similar degree (between
50--70 genes for each cell type), the vulD and vulC cells had the largest
enrichment effect size and the lowest q-values, suggesting that these cells
are more likely to be associated with an Egl phenotype than the others. This may
reflect the fact that vulD and vulE are the site of attachment for four vulval
muscles, vm1. Perhaps this attachment is particularly fragile, and perturbations
to these cells prevent adequate function of these muscles. In support of these
observations, P7.pa had the largest fold-enrichment of any tissue. In \cel{},
P7.pa gives rise to vulD and vulC. However, vulF is
also attached to a set of four additional vulval muscles, vm2. Why
is vulF less associated with an Egl phenotype?

@@

We calculated the conditional probability that a gene has an Egl phenotype given
that it's expression pattern includes a tissue term $X$ and we searched for the
tissue terms that maximized this probability. The list of terms that maximized this
probability reflected the results from running TEA on the subset of genes that have
an Egl phenotype. We also calculated the conditional probability that a gene has
expression in a tissue term $X$ given that it is annotated with an Egl phenotype
and we searched for terms that maximized this probability. The terms that maximized
this probability were `nervous system', `pharynx' (a body part with a lot of neurons),
`sex organ' and `tail' (a body part with neurons and hypodermis). In general, the
terms that had a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$.
Additionally, the terms that had a high $P(X|\text{Egl})$ are broad terms that
include a lot of cells, whereas the terms that had a high $P(\text{Egl}|X)$
were considerably more specific.
We conclude that the Egl phenotype arises from a small set of tissues. The Egl
phenotype can be best predicted by genes with expression patterns that include
at least one of a small number of cells (mainly vul cells, HSN). On the other
hand, answering whether the expression pattern of a gene includes a particular
anatomic region or tissue given that the gene has an Egl phenotype is hard to
do for small tissues or single cells. However, guesses about what functional system
or broad anatomic region may be affected by an Egl gene can be answered with
confidence ($\sim70\%$ of the time, the nervous system is the system
affected by an Egl mutant).

>>>>

% this is required to include graphics
\usepackage{graphicx}
% set the figures directory
\graphicspath{{./figs/}}

@@

\begin{figure}[htbp]
  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
  \centering
  \includegraphics[width=\linewidth]{gwas-design.pdf}
  \caption{Experimental design for human-nematode phenologue identification.
           We used GWAS candidates from the EBI-NHGRI catalog to identify
           disease associated genes, then used DIOPT to identify candidate
           homologs in \cel{}. Homologs were used to run phenotype, gene
           and tissue ontology enrichment analyses to identify disease
           phenologues.}
\label{fig:gwas}
\end{figure}

After we identified worm homologs for each trait, we reassessed how many traits
still had $>100$ gene candidates, and dropped all traits that had less than this
for our analysis. We identified 18 traits that met this criteria. The gene
lists for each of these 18 traits were then analyzed for gene, tissue and
phenotype enrichment (see~Fig.\ref{fig:gwas}). Tissue enrichment was
performed using the WormBase Tissue Enrichment Analysis (TEA) tool~\cite{}.

@@

Lupus-associated homologs were reasonably well annotated. Slightly more than half
of the genes had at least one phenotype annotation ($76/\wlupus{}$) and almost
all genes were annotated to at least one tissue or gene ontology term
($104/\wlupus{}$ and $115/\wlupus$ genes respectively).
We found that Lupus-associated homologs were enriched in `aneuploidy' (7 genes,
\qval{1}) and `meiotic chromosome segregation' (8 genes, \qval{1}). `Cell fate
transformation' (6 genes, \qval{1}), and `excess intestinal cells'
(5 genes, \qval{1}) were also overrepresented, as was `male tail morphology'
(6 genes, \qval{1}). Finally, the phenotype `nonsense mRNA accumulation' was also
enriched (5 genes, \qval{1}) (see Fig.~\ref{fig:lupus}).

\begin{figure}[htbp]
  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
  \centering
  \includegraphics[width=\linewidth]{systemic-lupus.pdf}
  \caption{Phenologue identification for systemic lupus erythematosus.
           \textbf{A} Phenotype Enrichment Analysis. \textbf{B} GO Enrichment
           Analysis. \textbf{C} Lupus in \cel{} may be best represented by
           a combination of three phenotypes: Cell proliferation possibly
           accompanied by aneuploidy; cell fate transformations that may lead
           to dysmorphias; and a molecular phenotype involving impairment of
           the Nonsense-Mediated Decay (NMD) pathway.}
\label{fig:lupus}
\end{figure}

TEA suggested that the
`excretory duct cell' (5 genes, \qval{2}) and the `posterior gonad arm' are
overrepresented in this dataset. We also found that the Pn.p cells P3.p through
P8.p were enriched in this dataset (5 genes, \qval{1}). GO enrichment pointed
at `modification-dependent macromolecule catabolic process' (23 genes,
\qval{15}) as a molecular function that characterizes this dataset. However,
this GO term was enriched only due to a single gene family, the \emph{skr} gene
family. Almost the entire \emph{skr} family was considered a candidate homolog
to the SKP1 human gene, making the GO enrichment suspect.

@@

\subsubsection*{Enrichment in the ciliary neuronal transcriptome}
As an example of how ontology enrichment can improve our understanding of
transcriptomes,
we selected a ciliary neuron dataset~\cite{} and ran the complete WormBase
Enrichment Suite on it. Ciliary neurons are present in the \cel{} male tail, but
they are also present in the male cephalic sensillum and hermaphrodites also have
ciliated neurons. PEA reveals that the ciliated
neuron transcriptome is enriched for genes that are typically associated with
`meiotic chromosome segregation' (46 genes, \qval{5}),
`aneuploidy' (42 genes, \qval{5}) and `spindle defective early embryos' (45
genes, \qval{2}) (see Fig.~\ref{fig:cilia}).

\begin{figure}[htbp]
  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
  \centering
  \includegraphics[width=\linewidth]{ciliary-transcriptome.pdf}
  \caption{PEA shows that the ciliary transcriptome is enriched in phenotypes
  related to cell division. Although this could reflect enrichment of
  microtubules and microtubule-related genes, the enrichment is at least
  partially driven by cell-cycle and DNA repair genes.}
\label{fig:cilia}
\end{figure}

@@

We extracted all of the \cel{} genes that have been associated with an Egl
phenotype and we used TEA to understand what tissues are enriched. The HSN
was enriched more than five-fold above background (\qval{7}) as were vulD, vulC,
vulE and vulF (\qval{6}). The vulA, vulB2 and vulB1 were
enriched at slightly lower levels (\qval{5}), whereas the uterine muscles and
uterine seam cells were enriched more than twice above background levels
(\qval{2}) (see Fig.~\ref{fig:egl}).
Therefore, the Egl phenotype would seem to emerge primarily from defects in the
HSN, secondarily from defects in the vulva, and only sometimes from defects in the
uterine seam cells or muscles. It is notable that all vulval cells were not equally
enriched. Although all the `vul' cells are annotated to a similar degree (between
50--70 genes for each cell type), the vulD and vulC cells had the largest
enrichment effect size and the lowest q-values, suggesting that these cells
are more likely to be associated with an Egl phenotype than the others. This may
reflect the fact that vulD and vulE are the site of attachment for four vulval
muscles, vm1. Perhaps this attachment is particularly fragile, and perturbations
to these cells prevent adequate function of these muscles. In support of these
observations, P7.pa had the largest fold-enrichment of any tissue. In \cel{},
P7.pa gives rise to vulD and vulC. However, vulF is
also attached to a set of four additional vulval muscles, vm2. Why
is vulF less associated with an Egl phenotype?

\begin{figure}[htbp]
  \renewcommand{\familydefault}{\sfdefault}\normalfont{}
  \centering
  \includegraphics[width=\linewidth]{Egl-phenotype.pdf}
  \caption{The Egl phenotype is a complex phenotype that is the result of
  interactions between many tissues. To dissect the contributions of individual
  tissues to generating an Egl phenotype, we obtained all genes annotated with
  with an Egl term. \textbf{A} We used TEA on the set of Egl genes to identify
  enriched tissues. \textbf{B} Anatomic diagram showing the tissues that are
  most enriched in the Egl gene set. Color coding shows the qualitative
  ordering of enrichment (red-Most enriched, yellow-least enriched). For clarity,
  not all cells are shown. All vm1 (4 cells) and vm2 (4 cells) muscles are
  symmetrical arranged around the vulva, but only 2 cells are shown for each.
  There are two seam cells on the left and right side of the vulva, but only
  cell on the right is shown. Only cells in the adult are shown.}
\label{fig:egl}
\end{figure}

@@

We calculated the conditional probability that a gene has an Egl phenotype given
that it's expression pattern includes a tissue term $X$ and we searched for the
tissue terms that maximized this probability. The list of terms that maximized this
probability reflected the results from running TEA on the subset of genes that have
an Egl phenotype. We also calculated the conditional probability that a gene has
expression in a tissue term $X$ given that it is annotated with an Egl phenotype
and we searched for terms that maximized this probability. The terms that maximized
this probability were `nervous system', `pharynx' (a body part with a lot of neurons),
`sex organ' and `tail' (a body part with neurons and hypodermis). In general, the
terms that had a high $P(\text{Egl}|X)$ did not have a high $P(X|\text{Egl})$.
Additionally, the terms that had a high $P(X|\text{Egl})$ are broad terms that
include a lot of cells, whereas the terms that had a high $P(\text{Egl}|X)$
were considerably more specific.
We conclude that the Egl phenotype arises from a small set of tissues. The Egl
phenotype can be best predicted by genes with expression patterns that include
at least one of a small number of cells (mainly vul cells, HSN). On the other
hand, answering whether the expression pattern of a gene includes a particular
anatomic region or tissue given that the gene has an Egl phenotype is hard to
do for small tissues or single cells. However, guesses about what functional system
or broad anatomic region an Egl gene is expressed in can be answered with
confidence ($\sim70\%$ of the time, an Egl mutant is expressed in
the nervous system).
